---
title-block-banner: true
title: "MMDBER_Physics Education"
format:
  html:
    toc: true
    toc-depth: 4 
    theme: Flatly
    page-layout: full
editor: visual
editor_options: 
  chunk_output_type: inline
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(haven)
library(glue)
library(MplusAutomation)
library(rhdf5)
library(here)
library(janitor)
library(gt)
library(semPlot)
library(reshape2)
library(cowplot)
library(filesstrings)
library(hrbrthemes)
library(poLCA)
library(naniar)
library(psych)
library(flextable)
library(officer)
conflicted::conflicts_prefer(flextable::compose)
```

```{r}
options(scipen=999, digits=4)
```

```{r}
support_data <- read.csv(here("k-means", "SocialSupport.csv"))
```

```{r}
vis_miss(support_data)
```

```{r}
describe(support_data)
```

```{r}
cor(support_data$Woman, support_data$Transgender)
cor(support_data$Woman, support_data$NonBinary)
cor(support_data$Transgender, support_data$NonBinary)
```

```{r}
conflicted::conflicts_prefer(dplyr::filter)
```

```{r}
nonbinary_subset <- support_data %>% filter(NonBinary == 1) #94 people
trans_subset <- support_data %>% filter(Transgender == 1)
nowoman_subset <- support_data %>% filter(Woman == 0) #66 people
nowoman_nonbinary <- nowoman_subset %>% filter(NonBinary == 1) #58 people
filtered_data <- support_data %>%
  filter(Woman == 1 | NonBinary == 1 | Transgender == 1)
```

## LCA

### Enumeration

```{r eval = FALSE}
lca_7  <- lapply(1:7, function(k) {
  lca_enum  <- mplusObject(
      
    TITLE = glue("{k}-Class"), 
  
    VARIABLE = glue(
    "categorical = Family-MentalHlth; 
     usevar = Family-MentalHlth;
     classes = c({k}); "),
  
  ANALYSIS = 
   "estimator = mlr; 
    type = mixture;
    starts = 200 100; 
    processors = 10;",
  
  OUTPUT = "sampstat residual tech11 tech14 svalues;",
  
  PLOT = 
    "type = plot3; 
    series = Family-MentalHlth(*);",
  
  usevariables = colnames(support_data),
  rdata = support_data)

lca_enum_fit <- mplusModeler(lca_enum, 
                            dataout=glue("support.dat"),
                            modelout=glue("c{k}_support.inp") ,
                            check=TRUE, run = TRUE, hashfilename = FALSE)
})
```

### Table of Fit

```{r}
source(here("k-means", "function/extract_mplus_info.R"))

output_dir <- here("k-means", "enum_mplus")

output_files <- list.files(output_dir, pattern = "\\.out$", full.names = TRUE)

final_data <- map_dfr(output_files, extract_mplus_info_extended)

sample_size <- unique(final_data$Sample_Size)
```

```{r warning=FALSE}
output <- readModels("k-means", "enum_mplus", filefilter = "support", quiet = TRUE)

enum_extract <- LatexSummaryTable(output,
    keepCols = c("Title","Parameters","LL","BIC","aBIC",
    "BLRT_PValue","T11_VLMR_PValue","Observations"),
    sortBy = "Title") 

allFit <- enum_extract %>%
  mutate(CAIC = -2 * LL + Parameters * (log(Observations) + 1)) %>%
  mutate(AWE = -2 * LL + 2 * Parameters * (log(Observations) + 1.5)) %>%
  mutate(SIC = -.5 * BIC) %>%
  mutate(expSIC = exp(SIC - max(SIC))) %>%
  mutate(BF = exp(SIC - lead(SIC))) %>%
  mutate(cmPk = expSIC / sum(expSIC)) %>%
  dplyr::select(1:5, 9:10, 6:7, 13, 14) %>%
  arrange(Parameters)

allFit <- allFit %>%
  mutate(Title = str_trim(Title)) %>%
  left_join(
    final_data %>%
      dplyr::select(Class_Model, Perc_Convergence, Replicated_LL_Perc, 
             Smallest_Class, Smallest_Class_Perc),
    by = c("Title" = "Class_Model")
  ) %>%
  mutate(Smallest_Class = coalesce(Smallest_Class, 
                                   final_data$Smallest_Class[match(Title, final_data$Class_Model)])) %>%
  relocate(Perc_Convergence, Replicated_LL_Perc, .after = LL) %>%
  mutate(Smallest_Class_Combined = paste0(Smallest_Class, "\u00A0(", Smallest_Class_Perc, "%)")) %>%  # ✅ Use Non-Breaking Space
  dplyr::select(-Smallest_Class, -Smallest_Class_Perc)  # ✅ Remove old columns to avoid duplication

allFit <- allFit %>%
  dplyr::select(
    Title, Parameters, LL,  # Core model fit stats
    Perc_Convergence, Replicated_LL_Perc,  # ✅ Move these next to LL
    BIC, aBIC, CAIC, AWE,  # Other fit indices
    T11_VLMR_PValue, BLRT_PValue,  # Fit test p-values
    Smallest_Class_Combined,  # ✅ Merged column in correct position
    BF, cmPk  # ✅ Comment these out in Flextable, but keep them in the data
  )
```

```{r}
fit_table1 <- allFit %>%
  dplyr::select(Title, Parameters, LL, Perc_Convergence, Replicated_LL_Perc, 
         BIC, aBIC, CAIC, AWE, 
         T11_VLMR_PValue, BLRT_PValue, 
         Smallest_Class_Combined  
         # BF,  # 🔹 Commented Out
         # cmPk  # 🔹 Commented Out
  ) %>%
  gt() %>%
  tab_header(title = md("**Model Fit Summary Table**")) %>%
  
  # ✅ Superheaders for groups of columns
  tab_spanner(label = "Model Fit Indices", columns = c(BIC, aBIC, CAIC, AWE)) %>%
  tab_spanner(label = "LRTs", columns = c(T11_VLMR_PValue, BLRT_PValue)) %>%
  tab_spanner(label = md("Smallest\u00A0Class"), columns = c(Smallest_Class_Combined)) %>%  # ✅ Non-Breaking Space
  
  cols_label(
    Title = "Classes",
    Parameters = md("npar"),
    LL = md("*LL*"),
    Perc_Convergence = "% Converged",
    Replicated_LL_Perc = "% Replicated",
    BIC = "BIC",
    aBIC = "aBIC",
    CAIC = "CAIC",
    AWE = "AWE",
    T11_VLMR_PValue = "VLMR",
    BLRT_PValue = "BLRT",
    Smallest_Class_Combined = "n (%)"
  ) %>%
  tab_footnote(
    footnote = md(
      "*Note.* Par = Parameters; *LL* = model log likelihood;
      BIC = Bayesian information criterion;
      aBIC = sample size adjusted BIC; CAIC = consistent Akaike information criterion;
      AWE = approximate weight of evidence criterion;
      BLRT = bootstrapped likelihood ratio test p-value;
      VLMR = Vuong-Lo-Mendell-Rubin adjusted likelihood ratio test p-value;
      *cmPk* = approximate correct model probability;
      Smallest K = Number of cases in the smallest class (n (%));
      LL Replicated = Whether the best log-likelihood was replicated."
    ),
    locations = cells_title()
  ) %>%
  tab_options(column_labels.font.weight = "bold") %>%
  
  # ✅ Format numerical values
  fmt_number(
    columns = c(3, 6:9),  # Keep decimal formatting for model fit statistics
    decimals = 2
  ) %>%
  fmt(
    columns = c(T11_VLMR_PValue, BLRT_PValue),  # ✅ Handle NA replacement & formatting
    fns = function(x) ifelse(is.na(x), "—", ifelse(x < 0.001, "<.001", scales::number(x, accuracy = .01)))
  ) %>%
  
  # ✅ Format as percentages (keeps 0-100 format)
  fmt_percent(
    columns = c(Perc_Convergence, Replicated_LL_Perc),
    decimals = 0,
    scale_values = FALSE  # ⚠️ IMPORTANT: Keeps numbers in 0-100 range
  ) %>%
  
  cols_align(align = "center", columns = everything()) %>%  # ✅ Center ALL columns
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = list(
      cells_body(columns = BIC, row = BIC == min(BIC)),
      cells_body(columns = aBIC, row = aBIC == min(aBIC)),
      cells_body(columns = CAIC, row = CAIC == min(CAIC)),
      cells_body(columns = AWE, row = AWE == min(AWE)),
      cells_body(columns = T11_VLMR_PValue, 
                 row = ifelse(T11_VLMR_PValue < .05 & lead(T11_VLMR_PValue) > .05, T11_VLMR_PValue < .05, NA)),
      cells_body(columns = BLRT_PValue, 
                 row = ifelse(BLRT_PValue < .05 & lead(BLRT_PValue) > .05, BLRT_PValue < .05, NA))
    )
  )

fit_table1
```

### Examine Convergence and Loglikelihood Replications

```{r}
source("function/extract_mplus_info.R")
output_dir <- here("enum_mplus")
output_files <- list.files(output_dir, pattern = "\\.out$", full.names = TRUE)
final_data <- map_dfr(output_files, extract_mplus_info_extended)
sample_size <- unique(final_data$Sample_Size)
```

```{r}
source("function/summary_table.R")
source("function/error_visualization.R")

summary_table <- create_flextable(final_data, sample_size)
summary_table

```

### Check for Loglikelihood Replication

```{r message=FALSE, warning=FALSE, results='hide'}
source("function/ll_replication_plots.R")

ll_replication_tables <- generate_ll_replication_plots(final_data)

walk(names(ll_replication_tables), function(name) {
  table <- ll_replication_tables[[name]]
  
  if (knitr::is_latex_output()) {
    img_path <- here("figures", paste0("ll_replication_table_", name, ".png"))
    gtsave(table, img_path, expand = 0, vwidth = 1200, vheight = 80 * nrow(table))
    cat(sprintf("\\includegraphics[width=5in]{%s}\n\n", img_path))
  } else {
    print(table) # Works fine in HTML
  }
})
```

```{r}
ll_replication_table_all <- source("function/ll_replication_processing.R", local = TRUE)$value
ll_replication_table_all
```

### Plot

#### Comparison plot 1-7 Class

```{r}
model_results <- data.frame()

for (i in 1:length(output)) {
  
  temp <- output[[i]]$parameters$probability.scale %>%                                       
    mutate(model = paste(i,"-Class Model"))                                                  
  
  model_results <- rbind(model_results, temp)
}

rm(temp)

compare_plot <-
  model_results %>%
  filter(category == 2) %>%
  dplyr::select(est, model, LatentClass, param) %>%
  mutate(param = as.factor(str_to_lower(param))) 

compare_plot$param <- fct_inorder(compare_plot$param)

ggplot(
  compare_plot,
  aes(
    x = param,
    y = est,
    color = LatentClass,
    shape = LatentClass,
    group = LatentClass,
    lty = LatentClass
  )
) +
  geom_point() + 
  geom_line() +
  scale_colour_viridis_d() +
  facet_wrap( ~ model, ncol = 2) +
  labs(title = "Support",
       x = " ", y = "Probability") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
                          axis.text.x = element_text(angle = -45, hjust = -.1))    
```

#### 3-Class Plot

```{r}
model_results_3 <- output[[3]]$parameters$probability.scale %>%
  mutate(model = "3 -Class Model")

plot_3 <-
  model_results_3 %>%
  filter(category == 2) %>%
  dplyr::select(est, model, LatentClass, param) %>%
  mutate(param = as.factor(str_to_lower(param))) 

plot_3$param <- fct_inorder(plot_3$param)

plot_3$param <- fct_recode(plot_3$param,
  "Family members" = "family",
  "Physics classmates/peers" = "physpeers",
  "Friends/Peers not in Physics" = "nonphyspee",
  "High school physics teacher(s)" = "hsphysteac",
  "Faculty (not physics) at your institution" = "nonphysfac",
  "Physics faculty at your institution" = "physfacult",
  "Research group/research mentors" = "researchgr",
  "Women in physics or STEM groups" = "wipgroup",
  "Gender inclusive physics/STEM groups" = "gipgroup",
  "Religious /faith based organizations" = "religorg",
  "Other identity-based affinity groups" = "otherident",
  "Social communities (e.g., book clubs, fitness group)" = "socialcomm",
  "Mental health professionals" = "mentalhlth"
)

plot_3$LatentClass <- recode(plot_3$LatentClass,
  `1` = "Class 1 (27.34%)",
  `2` = "Class 2 (49.73%)",
  `3` = "Class 3 (22.93%)"
)

ggplot(
  plot_3,
  aes(
    x = param,
    y = est,
    color = LatentClass,
    shape = LatentClass,
    group = LatentClass,
    lty = LatentClass
  )
) +
  geom_point() + 
  geom_line() +
  scale_colour_viridis_d() +
  facet_wrap( ~ model, ncol = 2) +
  labs(x = " ", y = "Probability of Item Endorsement") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
                          axis.text.x = element_text(angle = -30, hjust = -.1)) +
  labs(color = "Latent class", shape = "Latent class", lty = "Latent class")


```

#### 2-Class Plot

```{r}
model_results_2 <- output[[2]]$parameters$probability.scale %>%
  mutate(model = "2 -Class Model")

plot_2 <-
  model_results_2 %>%
  filter(category == 2) %>%
  dplyr::select(est, model, LatentClass, param) %>%
  mutate(param = as.factor(str_to_lower(param))) 

plot_2$param <- fct_inorder(plot_2$param)

plot_2$param <- fct_recode(plot_2$param,
  "Family members" = "family",
  "Physics classmates/peers" = "physpeers",
  "Friends/Peers not in Physics" = "nonphyspee",
  "High school physics teacher(s)" = "hsphysteac",
  "Faculty (not physics) at your institution" = "nonphysfac",
  "Physics faculty at your institution" = "physfacult",
  "Research group/research mentors" = "researchgr",
  "Women in physics or STEM groups" = "wipgroup",
  "Gender inclusive physics/STEM groups" = "gipgroup",
  "Religious /faith based organizations" = "religorg",
  "Other identity-based affinity groups" = "otherident",
  "Social communities (e.g., book clubs, fitness group)" = "socialcomm",
  "Mental health professionals" = "mentalhlth"
)

plot_2$LatentClass <- recode(plot_2$LatentClass,
  `1` = "Class 1 (73.90%)",
  `2` = "Class 2 (26.10%)"
)

ggplot(
  plot_2,
  aes(
    x = param,
    y = est,
    color = LatentClass,
    shape = LatentClass,
    group = LatentClass,
    lty = LatentClass
  )
) +
  geom_point() + 
  geom_line() +
  scale_colour_viridis_d() +
  facet_wrap( ~ model, ncol = 2) +
  labs(x = " ", y = "Probability of Item Endorsement") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
                          axis.text.x = element_text(angle = -30, hjust = -.1)) +
  labs(color = "Latent class", shape = "Latent class", lty = "Latent class")


```

### Observed Response Patterns

```{r message=FALSE, warning=FALSE, results='hide'}
patterns  <- mplusObject(
  
  TITLE = "LCA - Save response patterns", 
  
  VARIABLE = 
  "categorical = Family-MentalHlth; 
   usevar =  Family-MentalHlth;
   classes = c(3);",
  
  ANALYSIS = 
   "estimator = mlr; 
    type = mixture;    
    starts = 0; 
    processors = 10;
    optseed = 794236;",
  
  SAVEDATA = 
   "File=savedata.dat;
    Save=cprob;
    
    ! Code to save response frequency data 
    
    response is resp_patterns.dat;",
  
  OUTPUT = "patterns tech10 tech11 tech14",
  
  usevariables = colnames(support_data),
  rdata = support_data)

patterns_fit <- mplusModeler(patterns,
                dataout=("patterns.dat"),
                modelout=("patterns.inp") ,
                check=TRUE, run = TRUE, hashfilename = FALSE)
```

```{r warning=FALSE, message=FALSE}
# Read in response frequency data that we just created:
patterns <- read_table("resp_patterns.dat",
                        col_names=FALSE, na = "*") 

# Extract the column names
names <- names(readModels("patterns.out")[['savedata']])

# Add the names back to the dataset
colnames(patterns) <- c("Frequency", names)  
```

```{r}
# Order responses by highest frequency
order_highest <- patterns %>% 
  arrange(desc(Frequency)) 

# Loop `patterns` data to list top 5 conditional response patterns for each class
loop_cond  <- lapply(1:max(patterns$C), function(k) {       
order_cond <- patterns %>%                    
  filter(C == k) %>%                    
  arrange(desc(Frequency)) %>%                
  head(5)                                     
  })                                          

# Convert loop into data frame
table_data <- as.data.frame(bind_rows(loop_cond))

# Combine unconditional and conditional responses patterns
response_patterns <-  rbind(order_highest[1:5,], table_data) 
```

```{r}
resp_table <- response_patterns %>% 
  gt() %>%
    tab_header(
    title = "Observed Response Patterns",
    subtitle = html("Response patterns, estimated frequencies, estimated posterior class probabilities and modal assignments")) %>% 
    tab_source_note(
    source_note = md("Data Source: **Physics Education Data**")) %>%
    cols_label(
      Frequency = html("<i>f</i><sub>r</sub>"),
    CPROB1 = html("P<sub><i>k</i></sub>=1"),
    CPROB2 = html("P<sub><i>k</i></sub>=2"), 
    CPROB3 = html("P<sub><i>k</i></sub>=3"),# Change based on number of classes
    C = md("*k*")) %>% 
  tab_row_group(
    label = "Unconditional response patterns",
    rows = 1:5) %>%
  tab_row_group(
    label = md("*k* = 1 Conditional response patterns"),
    rows = 6:10) %>% #EDIT THESE VALUES BASED ON THE LAST COLUMN
  tab_row_group(
    label = md("*k* = 2 Conditional response patterns"),
    rows = 11:15)  %>% #EDIT THESE VALUES BASED ON THE LAST COLUMN
  tab_row_group(
    label = md("*k* = 3 Conditional response patterns"),
    rows = 16:20)  %>% #EDIT THESE VALUES BASED ON THE LAST COLUMN
    row_group_order(
      groups = c("Unconditional response patterns",
                 md("*k* = 1 Conditional response patterns"),
                 md("*k* = 2 Conditional response patterns"),
                 md("*k* = 3 Conditional response patterns"))) %>% 
    tab_footnote(
    footnote = html(
      "<i>Note.</i> <i>f</i><sub>r</sub> = response pattern frequency; P<sub><i>k</i></sub> = posterior class probabilities"
    )
  ) %>% 
  cols_align(align = "center") %>% 
  opt_align_table_header(align = "left") %>% 
  gt::tab_options(table.font.names = "Times New Roman") 

resp_table
```

### Classification Diagnostic Table

```{r message=FALSE, warning=FALSE, results='hide'}
classification  <- mplusObject(
  
  TITLE = "LCA - Calculated k-Class 95% CI",
  
  VARIABLE = 
  "categorical = Family-MentalHlth; 
   usevar =  Family-MentalHlth;
   classes = c(3);",
  
  ANALYSIS = 
   "estimator = ml; 
    type = mixture;    
    starts = 0; 
    processors = 10;
    optseed = 794236;
    bootstrap = 1000;",
  
  MODEL =
    "
  !CHANGE THIS SECTION TO YOUR CHOSEN k-CLASS MODEL
    
  %OVERALL%
  [C#1](c1);
  [C#2](c2);

  Model Constraint:
  New(p1 p2 p3);
  
  p1 = exp(c1)/(1+exp(c1)+exp(c2));
  p2 = exp(c2)/(1+exp(c1)+exp(c2));
  p3 = 1/(1+exp(c1)+exp(c2));",

  
  OUTPUT = "cinterval(bcbootstrap)",
  
  usevariables = colnames(support_data),
  rdata = support_data)

classification_fit <- mplusModeler(classification,
                dataout=("class.dat"),
                modelout=("class.inp") ,
                check=TRUE, run = TRUE, hashfilename = FALSE)
```

```{r warning=FALSE}
class_output <- readModels("class.out")

entropy <- c(class_output$summaries$Entropy, rep(NA, class_output$summaries$NLatentClasses-1))
entropy

k_ci <- class_output$parameters$ci.unstandardized %>% 
  filter(paramHeader == "New.Additional.Parameters") %>% 
  unite(CI, c(low2.5,up2.5), sep=", ", remove = TRUE) %>% 
  mutate(CI = paste0("[", CI, "]")) %>% 
  rename(kclass=est) %>% 
  dplyr::select(kclass, CI)

avePPk <- tibble(avePPk = diag(class_output$class_counts$avgProbs.mostLikely))

mcaPk <- round(class_output$class_counts$mostLikely,3) %>% 
  mutate(model = paste0("Class ", class)) %>% 
  add_column(avePPk, k_ci) %>% 
  rename(mcaPk = proportion) %>% 
  dplyr::select(model, kclass, CI, mcaPk, avePPk)

OCCk <- mcaPk %>% 
  mutate(OCCk = round((avePPk/(1-avePPk))/(kclass/(1-kclass)),3))

class_df <- data.frame(OCCk)

class_table <- class_df %>% 
  gt() %>%
    tab_header(
    title = "Model Classification Diagnostics for the 3-Class Solution") %>%
    cols_label(
      model = md("*k*-Class"),
      kclass = md("*k*-Class Proportions"),
      CI = "95% CI",
      mcaPk = md("*mcaPk*"),
      avePPk = md("*AvePPk*"),
      OCCk = md("*OCCk*")) %>% 
    sub_missing(6,
              missing_text = "") %>%
  cols_align(align = "center") %>% 
  opt_align_table_header(align = "left") %>% 
  gt::tab_options(table.font.names = "Times New Roman")

class_table
```

### Step 1 - Class Enumeration w/ Auxiliary Specification

```{r message=FALSE, warning=FALSE, results='hide'}
step1  <- mplusObject(
  TITLE = "Step 1", 
  VARIABLE = 
  "categorical = Family-MentalHlth; 
   usevar = Family-MentalHlth;
    
   classes = c(3); 
    
   auxiliary =   ! list all potential covariates and distals here
   Woman NonBinary Transgender     ! covariate
   Identity;      ! distal outcome ",
  
  ANALYSIS = 
   "estimator = mlr; 
    type = mixture;
    starts = 0;
    optseed = 794236;",
  
  SAVEDATA = 
   "File=3step_savedata.dat;
    Save=cprob;",
  
  OUTPUT = "residual tech11 tech14",
  
  PLOT = 
    "type = plot3; 
    series = Family-MentalHlth(*);",
  
  usevariables = colnames(support_data),
  rdata = support_data)

step1_fit <- mplusModeler(step1,
                            dataout=("Step1.dat"),
                            modelout=("Step1.inp") ,
                            check=TRUE, run = TRUE, hashfilename = FALSE)
```

```{r eval=FALSE, message=FALSE, warning=FALSE, results='hide'}
#for 2 class modal assignment only
step1_2class  <- mplusObject(
  TITLE = "Step 1_2class", 
  VARIABLE = 
  "categorical = Family-MentalHlth; 
   usevar = Family-MentalHlth;
    
   classes = c(2); 
    
   auxiliary =   ! list all potential covariates and distals here
   Woman NonBinary Transgender     ! covariate
   Identity;      ! distal outcome ",
  
  ANALYSIS = 
   "estimator = mlr; 
    type = mixture;
    starts = 0;
    optseed = 987090;",
  
  SAVEDATA = 
   "File=3step_savedata.dat;
    Save=cprob;",
  
  OUTPUT = "residual tech11 tech14",
  
  PLOT = 
    "type = plot3; 
    series = Family-MentalHlth(*);",
  
  usevariables = colnames(support_data),
  rdata = support_data)

step1_fit_2class <- mplusModeler(step1_2class,
                            dataout=("Step1_2class.dat"),
                            modelout=("Step1_2class.inp") ,
                            check=TRUE, run = TRUE, hashfilename = FALSE)
```

```{r}
source("function/plot_lca.txt")
output_lca <- readModels("Step1.out")

plot_lca(model_name = output_lca)
```

### Step 2 - Determine Measurement Error

```{r}
logit_cprobs <- as.data.frame(output_lca[["class_counts"]]
                                       [["logitProbs.mostLikely"]])

savedata <- as.data.frame(output_lca[["savedata"]])

colnames(savedata)[colnames(savedata)=="C"] <- "N"
```

### Step 3 - Add Auxiliary Variables (Non-binary status and physics identity)

```{r message=FALSE, warning=FALSE}
step3  <- mplusObject(
  TITLE = "Step3;", 
  
  VARIABLE = 
 "nominal = N;
  usevar = n;
  
  classes = c(3);
  
  usevar = NonBinar Identity;" ,
  
  ANALYSIS = 
 "estimator = mlr; 
  type = mixture; 
  starts = 0;",
 
  DEFINE = 
   "center NonBinar (grandmean);",
  
  MODEL =
  glue(
 " %OVERALL%
 
  Identity on NonBinar; 
  C on NonBinar;  

  %C#1%
  [n#1@{logit_cprobs[1,1]}];
  [n#2@{logit_cprobs[1,2]}];
  
  [Identity](m1);   
  Identity;    

  %C#2%
  [n#1@{logit_cprobs[2,1]}];
  [n#2@{logit_cprobs[2,2]}];
  
  [Identity](m2);
  Identity;
  
  %C#3%
  [n#1@{logit_cprobs[3,1]}];
  [n#2@{logit_cprobs[3,2]}];
  
  [Identity](m3);
  Identity;"),
  
  MODELCONSTRAINT = 
   "New (diff12 diff13 diff23);
  
    diff12 = m1-m2; 
    diff13 = m1-m3;
    diff23 = m2-m3;",
  
  MODELTEST = "   
    m1=m2;
    m2=m3;",
 
  usevariables = colnames(savedata), 
  rdata = savedata)

step3_fit <- mplusModeler(step3,
               dataout=("Step3.dat"), 
               modelout=("Step3.inp"), 
               check=TRUE, run = TRUE, hashfilename = FALSE)
```

```{r}
modelParams <- readModels("Step3.out")

wald <- as.data.frame(modelParams[["summaries"]]) %>%
  dplyr::select(WaldChiSq_Value:WaldChiSq_PValue) %>% 
  mutate(WaldChiSq_DF = paste0("(", WaldChiSq_DF, ")")) %>% 
  unite(wald_test, WaldChiSq_Value, WaldChiSq_DF, sep = " ") %>% 
  rename(pval = WaldChiSq_PValue) %>% 
  mutate(pval = ifelse(pval<0.001, paste0("<.001*"),
                       ifelse(pval<0.05, paste0(scales::number(pval, accuracy = .001), "*"),
                              scales::number(pval, accuracy = .001))))

wald_table <- wald %>% 
  gt() %>%
    tab_header(
    title = "Wald Test of Paramter Constraints (Math)") %>%
    cols_label(
      wald_test = md("Wald Test (*df*)"),
      pval = md("*p*-value")) %>% 
  cols_align(align = "center") %>% 
  opt_align_table_header(align = "left") %>% 
  gt::tab_options(table.font.names = "serif")

wald_table
```

```{r}
# Extract information as data frame
diff <- as.data.frame(modelParams[["parameters"]][["unstandardized"]]) %>%
  filter(grepl("DIFF", param)) %>% 
  dplyr::select(param:pval) %>% 
  mutate(se = paste0("(", format(round(se,2), nsmall =2), ")")) %>% 
  unite(estimate, est, se, sep = " ") %>% 
  mutate(param = str_remove(param, "DIFF"),
         param = as.numeric(param)) %>% 
  separate(param, into = paste0("Group", 1:2), sep = 1) %>% 
  mutate(class = paste0("Class ", Group1, " vs ", Group2)) %>% 
  dplyr::select(class, estimate, pval)

# Create table

diff %>% 
  gt() %>%
    tab_header(
    title = "Distal Outcome Differences") %>%
    cols_label(
      class = "Class",
      estimate = md("Mean (*se*)"),
      pval = md("*p*-value")) %>% 
    sub_missing(1:3,
              missing_text = "") %>%
  fmt(3,
    fns = function(x)
      ifelse(x<0.05, paste0(scales::number(x, accuracy = .001), "*"),
             scales::number(x, accuracy = .001))
  ) %>% 
  cols_align(align = "center") %>% 
  opt_align_table_header(align = "left") %>% 
  gt::tab_options(table.font.names = "serif")
```

## K-modes clustering

```{r message=FALSE, warning=FALSE}
library(tidymodels)
library(tidyverse)
library(modeldata)
library(kernlab)
library(tidyclust)
library(corrplot)
library(klaR)
library(cluster)
library(clValid)
library(caret)
library(lattice)
library(mclust)
library(Rtsne)
library(gmodels)
library(ggplot2)
library(klaR)
library(cluster) 
library(car)
tidymodels_prefer()
```

### Choose the optimal number of cluster

```{r eval=FALSE}
max_k <- 10

costs <- numeric(max_k)

df_selected <- savedata[, c("FAMILY", "PHYSPEER", "NONPHYSP", "HSPHYSTE", "NONPHYSF",
                            "PHYSFACU", "RESEARCH", "WIPGROUP", "GIPGROUP", "RELIGORG",
                            "OTHERIDE", "SOCIALCO", "MENTALHL")]  
```

#### Elbow plot

```{r eval=FALSE}
for(k in 1:max_k) {
  set.seed(123)  
  km <- kmodes(df_selected, modes = k, iter.max = 300)
  costs[k] <- sum(km$withindiff)
}

plot(1:max_k, costs, type="b", pch=19,
     xlab="Number of Clusters (k)",
     ylab="Total Within-Cluster Dissimilarity",
     main="Elbow Plot for K-Modes Clustering")
```

#### Silhouette plot

```{r eval=FALSE}
#specify silhouette width
diss_matrix <- daisy(df_selected, metric = "gower", type = list(symm = 1:ncol(df_selected)))
dist_matrix <- as.dist(diss_matrix)

avg_sil_width <- numeric(max_k) 
avg_sil_width[1] <- NA   

for(k in 2:max_k) {
  set.seed(123)
  km <- kmodes(df_selected, modes = k, iter.max = 300)
  cluster_labels <- km$cluster  
  costs[k] <- sum(km$withindiff)
  sil <- cluster::silhouette(cluster_labels, dist_matrix)
  avg_sil_width[k] <- mean(sil[, "sil_width"])  # average silhouette width
}

plot(2:max_k, avg_sil_width[2:max_k], type="b", pch=19,
     xlab="Number of Clusters (k)",
     ylab="Average Silhouette Width",
     main="Silhouette Analysis for k-modes")

avg_sil_width[2]
avg_sil_width[3]

optimal_k_sil <- which.max(avg_sil_width[2:max_k]) + 1 
optimal_k_sil
```

### 2-Cluster Solution

```{r}
num_clusters <- 2

df_2cluster <- savedata[, c("FAMILY", "PHYSPEER", "NONPHYSP", "HSPHYSTE", "NONPHYSF",
                            "PHYSFACU", "RESEARCH", "WIPGROUP", "GIPGROUP", "RELIGORG",
                            "OTHERIDE", "SOCIALCO", "MENTALHL")]  

set.seed(123) 
kmodes_result_2 <- kmodes(df_2cluster, num_clusters, iter.max = 300, weighted = FALSE)

df_2cluster$Cluster <- as.factor(kmodes_result_2$cluster)
```

#### Response patterns - mode

```{r}
print(kmodes_result_2$modes)
```

#### Response patterns plot

```{r}
cluster_totals <- df_2cluster %>%
  group_by(Cluster) %>%
  summarise(Total = n())


df_long <- df_2cluster %>%
  pivot_longer(
    cols = -Cluster,
    names_to = "Item",
    values_to = "Endorsed"
  )

endorsed_counts <- df_long %>%
  filter(Endorsed == 1) %>%
  group_by(Cluster, Item) %>%
  summarise(Count = n(), .groups = "drop") %>%
  left_join(cluster_totals, by = "Cluster") %>%
  mutate(Percentage = (Count / Total) * 100)

endorsed_counts$Item <- factor(endorsed_counts$Item)

endorsed_counts$Cluster <- dplyr::recode(endorsed_counts$Cluster,
  `1` = "Cluster 1",
  `2` = "Cluster 2"
)

endorsed_counts$Item <- dplyr::recode(tolower(endorsed_counts$Item),
  "family"       = "Family members",
  "physpeer"     = "Physics classmates/peers",
  "nonphysp"     = "Friends/Peers not in Physics",
  "hsphyste"     = "High school physics teacher(s)",
  "nonphysf"     = "Faculty (not physics) at your institution",
  "physfacu"     = "Physics faculty at your institution",
  "research"     = "Research group/research mentors",
  "wipgroup"     = "Women in physics or STEM groups",
  "gipgroup"     = "Gender inclusive physics/STEM groups",
  "religorg"     = "Religious /faith based organizations",
  "otheride"     = "Other identity-based affinity groups",
  "socialco"     = "Social communities (e.g., book clubs, fitness group)",
  "mentalhl"     = "Mental health professionals"
)

item_order <- c("Family members", "Physics classmates/peers", 
                "Friends/Peers not in Physics", "High school physics teacher(s)", 
                "Faculty (not physics) at your institution",
                "Physics faculty at your institution", "Research group/research mentors",
                "Women in physics or STEM groups", "Gender inclusive physics/STEM groups",
                "Religious /faith based organizations",
                "Other identity-based affinity groups", 
                "Social communities (e.g., book clubs, fitness group)", 
                "Mental health professionals")
```

```{r}
plot_2cluster <- ggplot(endorsed_counts, aes(x = Percentage, y = Item, fill = Cluster)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")),
            position = position_dodge(width = 0.7),
            hjust = ifelse(endorsed_counts$Percentage < 10, -0.1, -0.2), 
            size = 2.5) +
  scale_y_discrete(limits = rev(item_order)) +
  theme_minimal() +
  labs(
    title = "Item Endorsements by Cluster",
    x = "Percentage of Individuals",
    y = "Item Endorsement"
  ) +
  theme(
    axis.text.y = element_text(size = 10),
    strip.text = element_text(face = "bold"),
    plot.margin = margin(10, 10, 10, 10),
    legend.position = "right",
    clip = "off"
  )

plot_2cluster
```

#### Cross comparison of K-modes clustering and LCA results - 2 clusters

```{r}
savedata$Cluster_2 <- as.factor(kmodes_result_2$cluster)

CrossTable(savedata$N, savedata$Cluster_2, prop.chisq = FALSE, prop.t = FALSE, 
           prop.r = TRUE, prop.c = TRUE, dnn = c("LCA Enumeration", "K-Modes Clustering"))
```

#### Comparison bar plot - 2 cluster/3 class

```{r}
cross_tab <- table(savedata$N, savedata$Cluster_2)
cross_tab

cross_tab_df <- as.data.frame(cross_tab)
colnames(cross_tab_df) <- c("LCA_Class", "KModes_Cluster", "Count")

cross_tab_df$LCA_Class <- as.factor(cross_tab_df$LCA_Class)
cross_tab_df$KModes_Cluster <- as.factor(cross_tab_df$KModes_Cluster)

cross_tab_df <- cross_tab_df %>%
  group_by(LCA_Class) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)

ggplot(cross_tab_df, aes(x = KModes_Cluster, y = Percentage, fill = KModes_Cluster)) +
  geom_bar(stat = "identity") +  
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), vjust = -0.5, size = 3) + 
  facet_wrap(~ LCA_Class, ncol = 3,  
             labeller = labeller(LCA_Class = c("1" = "LCA Class 1",
                                               "2" = "LCA Class 2",
                                               "3" = "LCA Class 3"))) +
  theme_minimal() +
  labs(y = "Proportion (%)",
       x = "K-Modes Cluster",
       fill = "K-Modes Cluster") +
  theme(legend.position = "bottom", 
        plot.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 10),
        strip.text = element_text(size = 10))
```

#### Comparison bar plot - 2 cluster/2 class

```{r}
output_2class <- readModels("Step1_2class.out")

logit_cprobs2 <- as.data.frame(output_2class[["class_counts"]]
                                       [["logitProbs.mostLikely"]])

savedata2 <- as.data.frame(output_2class[["savedata"]])

colnames(savedata2)[colnames(savedata2)=="C"] <- "N"
```

```{r}
cross_tab <- table(savedata2$N, savedata$Cluster_2)
cross_tab

cross_tab_df <- as.data.frame(cross_tab)
colnames(cross_tab_df) <- c("LCA_Class", "KModes_Cluster", "Count")

cross_tab_df$LCA_Class <- as.factor(cross_tab_df$LCA_Class)
cross_tab_df$KModes_Cluster <- as.factor(cross_tab_df$KModes_Cluster)

cross_tab_df <- cross_tab_df %>%
  group_by(LCA_Class) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)

ggplot(cross_tab_df, aes(x = KModes_Cluster, y = Percentage, fill = KModes_Cluster)) +
  geom_bar(stat = "identity") +  
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), vjust = -0.5, size = 3) + 
  facet_wrap(~ LCA_Class, ncol = 3,  
             labeller = labeller(LCA_Class = c("1" = "LCA Class 1",
                                               "2" = "LCA Class 2",
                                               "3" = "LCA Class 3"))) +
  theme_minimal() +
  labs(y = "Proportion (%)",
       x = "K-Modes Cluster",
       fill = "K-Modes Cluster") +
  theme(legend.position = "bottom", 
        plot.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 10),
        strip.text = element_text(size = 10))
```

#### Identity by K-modes cluster - 2 clusters

```{r}
summary_table <- savedata %>%
  group_by(Cluster_2) %>%
  summarise(Mean_Identity = mean(IDENTITY, na.rm = TRUE),  
            SD_Identity = sd(IDENTITY, na.rm = TRUE),     
            Count = n()) 

print(summary_table) 
```

#### ANCOVA - Relating cluster membership to physics identity

```{r}
model_ancova <- lm(IDENTITY ~ Cluster_2 + NONBINAR, data = savedata)
summary(model_ancova)

Anova(model_ancova, type = "III")

interaction_model <- lm(IDENTITY ~ Cluster_2 * NONBINAR, data = savedata)
summary(interaction_model)
```

#### Logistic regression & Chi-square - Relating cluster membership to non-binary status

```{r}
log_model <- glm(Cluster_2 ~ NONBINAR, data = savedata, family = "binomial")

summary(log_model)
```

```{r}
table_data <- table(savedata$Cluster_2, savedata$NONBINAR)
stats::chisq.test(table_data)
```

### 3-Cluster Solution

```{r}
num_clusters <- 3

df_3cluster <- savedata[, c("FAMILY", "PHYSPEER", "NONPHYSP", "HSPHYSTE", "NONPHYSF",
                            "PHYSFACU", "RESEARCH", "WIPGROUP", "GIPGROUP", "RELIGORG",
                            "OTHERIDE", "SOCIALCO", "MENTALHL")]  

set.seed(57) 
kmodes_result_3 <- kmodes(df_3cluster, num_clusters, iter.max = 200, weighted = FALSE)

df_3cluster$Cluster <- as.factor(kmodes_result_3$cluster)
```

#### Response patterns

```{r}
print(kmodes_result_3$modes)
```

#### Response patterns plot

```{r}
cluster_totals <- df_3cluster %>%
  group_by(Cluster) %>%
  summarise(Total = n())


df_long <- df_3cluster %>%
  pivot_longer(
    cols = -Cluster,
    names_to = "Item",
    values_to = "Endorsed"
  )

endorsed_counts <- df_long %>%
  filter(Endorsed == 1) %>%
  group_by(Cluster, Item) %>%
  summarise(Count = n(), .groups = "drop") %>%
  left_join(cluster_totals, by = "Cluster") %>%
  mutate(Percentage = (Count / Total) * 100)

endorsed_counts$Item <- factor(endorsed_counts$Item)

endorsed_counts$Cluster <- dplyr::recode(endorsed_counts$Cluster,
  `1` = "Cluster 1",
  `2` = "Cluster 2",
  `3` = "Cluster 3"
)

endorsed_counts$Item <- dplyr::recode(tolower(endorsed_counts$Item),
  "family"       = "Family members",
  "physpeer"     = "Physics classmates/peers",
  "nonphysp"     = "Friends/Peers not in Physics",
  "hsphyste"     = "High school physics teacher(s)",
  "nonphysf"     = "Faculty (not physics) at your institution",
  "physfacu"     = "Physics faculty at your institution",
  "research"     = "Research group/research mentors",
  "wipgroup"     = "Women in physics or STEM groups",
  "gipgroup"     = "Gender inclusive physics/STEM groups",
  "religorg"     = "Religious /faith based organizations",
  "otheride"     = "Other identity-based affinity groups",
  "socialco"     = "Social communities (e.g., book clubs, fitness group)",
  "mentalhl"     = "Mental health professionals"
)

item_order <- c("Family members", "Physics classmates/peers", 
                "Friends/Peers not in Physics", "High school physics teacher(s)", 
                "Faculty (not physics) at your institution",
                "Physics faculty at your institution", "Research group/research mentors",
                "Women in physics or STEM groups", "Gender inclusive physics/STEM groups",
                "Religious /faith based organizations",
                "Other identity-based affinity groups", 
                "Social communities (e.g., book clubs, fitness group)", 
                "Mental health professionals")

plot_3cluster <- ggplot(endorsed_counts, aes(x = Percentage, y = Item, fill = Cluster)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")),
            position = position_dodge(width = 0.7),
            hjust = ifelse(endorsed_counts$Percentage < 10, -0.1, -0.2), 
            size = 2.5) +
  scale_y_discrete(limits = rev(item_order)) +
  theme_minimal() +
  labs(
    title = "Item Endorsements by Cluster",
    x = "Percentage of Individuals",
    y = "Item Endorsement"
  ) +
  theme(
    axis.text.y = element_text(size = 10),
    strip.text = element_text(face = "bold"),
    plot.margin = margin(10, 10, 10, 10),
    legend.position = "right",
    clip = "off"
  )

plot_3cluster
```

#### Cross comparison of K-modes clustering and LCA results - 3 clusters

```{r}
savedata$Cluster_3 <- as.factor(kmodes_result_3$cluster)

CrossTable(savedata$N, savedata$Cluster_3, prop.chisq = FALSE, prop.t = FALSE, 
           prop.r = TRUE, prop.c = TRUE, dnn = c("LCA Enumeration", "K-Modes Clustering"))
```

#### Comparison bar plot

```{r}
cross_tab <- table(savedata$N, savedata$Cluster_3)
cross_tab

cross_tab_df <- as.data.frame(cross_tab)
colnames(cross_tab_df) <- c("LCA_Class", "KModes_Cluster", "Count")

cross_tab_df$LCA_Class <- as.factor(cross_tab_df$LCA_Class)
cross_tab_df$KModes_Cluster <- as.factor(cross_tab_df$KModes_Cluster)

cross_tab_df <- cross_tab_df %>%
  group_by(LCA_Class) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)

ggplot(cross_tab_df, aes(x = KModes_Cluster, y = Percentage, fill = KModes_Cluster)) +
  geom_bar(stat = "identity") +  
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), vjust = -0.5, size = 3) + 
  facet_wrap(~ LCA_Class, ncol = 3,  
             labeller = labeller(LCA_Class = c("1" = "LCA Class 1",
                                               "2" = "LCA Class 2",
                                               "3" = "LCA Class 3"))) +
  theme_minimal() +
  labs(y = "Proportion (%)",
       x = "K-Modes Cluster",
       fill = "K-Modes Cluster") +
  theme(legend.position = "bottom", 
        plot.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 10),
        strip.text = element_text(size = 10))
```
