
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE) 
```


# Comparing Cluster Analysis and  Latent Profile Analysis (B. J. Yik,* Y. Zhang, K. Nylund-Gibson, M. Ing, L. Krawiec, J. D. Houck, & E. D. Nacsa, 2026)

Citation: [Exploring Heterogeneity in Chemistry Education Research: Comparing Cluster Analysis and  Latent Profile Analysis (B. J. Yik,* Y. Zhang, K. Nylund-Gibson, M. Ing, L. Krawiec, J. D. Houck, & E. D. Nacsa (2026)](https://doi.org/10.1039/D5RP00432B)

```{r}
library(tidyverse)
library(haven)
library(glue)
library(MplusAutomation)
library(rhdf5)
library(here)
library(janitor)
library(semPlot)
library(reshape2)
library(cowplot)
library(filesstrings)
library(hrbrthemes)
library(sjPlot)
library(dplyr)
library(naniar)
library(gt)
library(tidyLPA)
#library(pisaUSA15)
library(patchwork)
library(RcppAlgos)
library(DiagrammeR)
library(data.table)
library(poLCA)
library(cluster)
library(factoextra)
library(datasets)
library(Hmisc)
library(car)
```

Proposed model applied in current study
```{r}
grViz(" digraph lca_model {

# The `graph` statement - No editing needed

    graph [layout = dot, overlap = true]
 
# Two `node` statements
 
# One for measured variables (box) 

    node [shape=box]
    Attitudes [label = <Attitudes toward <br/>chemistry>]
    Motivation
    Effort [label = <Effort Beliefs>]
    Efficacy [label = <Self-efficacy>] 
    Concept [label = <Self-concept>]
    Grading [label = <Grading Scheme>]
    Exam [label = <ACS Exam Score>];
 
# One for latent variables (circle) 
 
    node [shape=circle]
    dispositions [label=<Student <br/>Affects <br/>C<sub>k=3</sub>>];
    
# `edge` statements
 
    edge [minlen = 2]
    dispositions -> {Attitudes Motivation Effort Efficacy Concept}
    dispositions -> Exam [minlen = 4];
    Grading -> dispositions [minlen = 4];
    Grading -> Exam;
    
 {rank = same; Grading; dispositions; Exam}
 #{rank = source; Attitudes; Motivation; Effort; Efficacy; Concept}
 }") 
```

load the dataset
```{r}
#load data
pre <- read.csv(here("33-k-means", "grading-pre-v2.csv"))
```

handling missing data
```{r}
#any(is.na(pre))
pre_clean <- pre %>% drop_na()
any(is.na(pre_clean))

#we also cleaned out this few datapoint, since we previously included race as an auxiliary variable, but deleted after; but we did not change the sample for LPA. only 6 sample was deleted due to this step 
pre_clean <- pre_clean %>%
  filter(!Gender %in% c("I am gender nonconforming,I am genderqueer or genderfluid", "My gender or gender identity is best described as:", "I am gender nonconforming", "I prefer not to disclose my gender or gender identity", "I am genderqueer or genderfluid", "I am nonbinary" ))
```

view data
```{r, include = FALSE}
sjPlot::view_df(pre_clean)
```

recode
```{r}
pre_clean <- pre_clean %>%
  mutate(grading = ifelse(grading == "specs", 1, 0))  # specs = 1, trad = 0
```

calculate means for each scale
```{r}
pre_clean <- pre_clean %>%
  mutate(ASCI_IA = rowMeans(across(c(ASCI1:ASCI8)), na.rm = TRUE)) %>%
  mutate(AMSC = rowMeans(across(c(AMSC01:AMSC28)), na.rm = TRUE)) %>%
  mutate(EB = rowMeans(across(c(EBN1:EBP4)), na.rm = TRUE)) %>%
  mutate(MSLQ = rowMeans(across(c(MSLQ05:MSLQ31)), na.rm = TRUE)) %>%
  mutate(CSC = rowMeans(across(c(CSC24:CSC08)), na.rm = TRUE))

#update the data preview
sjPlot::view_df(pre_clean)
```

standardize the variables (z-score)
```{r}
pre_clean <- pre_clean %>%
  mutate(across(all_of(c("ASCI_IA", "AMSC", "EB", "MSLQ", "CSC")), ~ scale(.)[,1], .names = "z_{.col}"))
```

double check all indicators are standardized
```{r}
ds <- pre_clean %>%
    pivot_longer(z_ASCI_IA:z_CSC, names_to = "variable") %>%
    group_by(variable) %>%
    summarise(mean = mean(value, na.rm = TRUE), sd = sd(value, na.rm = TRUE))

ds %>%
    gt() %>%
    tab_header(title = md("**Descriptive Summary**")) %>%
    cols_label(variable = "Variable", mean = md("M"), sd = md("SD")) %>%
    fmt_number(c(2:3), decimals = 2) %>%
    cols_align(align = "center", columns = mean)

data_long <- pre_clean %>%
    pivot_longer(z_ASCI_IA:z_CSC, names_to = "variable")

ggplot(data_long, aes(x = value)) + geom_histogram(binwidth = 0.3, fill = "#69b3a2",
    color = "black") + facet_wrap(~variable, scales = "free_x") + labs(title = "Histograms of Variables",
    x = "Value", y = "Frequency") + theme_cowplot()
```
correlations
```{r}
df_cor <- pre_clean %>%
  dplyr::select(z_ASCI_IA, z_AMSC, z_EB, z_MSLQ, z_CSC, grading, exam)

cor_matrix <- cor(df_cor, use = "pairwise.complete.obs")

print(cor_matrix)

cor_result <- rcorr(as.matrix(df_cor))

formatted_p <- formatC(cor_result$P, format = "f", digits = 2)

dim(formatted_p) <- dim(cor_result$P)
rownames(formatted_p) <- colnames(formatted_p) <- colnames(df_cor)

print(formatted_p, quote = FALSE)
```


run LPA model
```{r, eval = FALSE}
# Run LPA models (no need to run it everytime)
lpa_fit <- pre_clean %>%
    dplyr::select(z_ASCI_IA, z_AMSC, z_EB, z_MSLQ, z_CSC) %>%
    estimate_profiles(1:6, package = "MplusAutomation", ANALYSIS = "starts = 500 100;",
        OUTPUT = "sampstat residual tech11 tech14", variances = c("equal", "varying",
            "equal", "varying"), covariances = c("zero", "zero",
            "equal", "varying"), keepfiles = TRUE)

#compare fit statistics
get_fit(lpa_fit)

# Move files to folder
files <- list.files(here(), pattern = "^model")
move_files(files, here("33-k-means", "new"), overwrite = TRUE)
```

```{r}
source(here("33-k-means", "functions", "enum_table_lpa.r")) #file in the folder

# Read in model
output_pisa <- readModels(here("33-k-means", "new"), quiet = TRUE)
```

```{r, eval = FALSE}
# Preview with numbered rows
enum_fit(output_pisa)
```

model fit summary table (this is the old version table, still working on the new one; the new one will include smallest profile)
```{r}
select_models <-LatexSummaryTable(output_pisa,                                 
                keepCols=c("Title", "Parameters", "LL", "BIC", "aBIC",
                           "BLRT_PValue", "T11_VLMR_PValue","Observations"))
```

```{r, eval = FALSE}
enum_table(select_models, 1:6, 7:12, 13:18, 19:23)
```

```{r, eval = FALSE}
source(here("33-k-means", "functions","ic_plot_lpa.R")) 
#file in the folder
ic_plot(output_pisa)
```

second round comparison among different profiles
```{r}
# CmpK recalculation:
enum_fit1 <- select_models

stage2_cmpk <- enum_fit1 %>% 
  slice(2, 3, 6, 8, 12, 16, 20, 22) %>% # Change this to select the rows of the candidate models
  mutate(CAIC = -2 * LL + Parameters * (log(Observations) + 1)) %>%
  mutate(AWE = -2 * LL + 2 * Parameters * (log(Observations) + 1.5)) %>%
  mutate(SIC = -.5 * BIC,
         expSIC = exp(SIC - max(SIC)),
         cmPk = expSIC / sum(expSIC),
         BF = exp(SIC - lead(SIC)))  %>%
  dplyr::select(Title, Parameters, BIC, aBIC, CAIC, AWE, cmPk, BF) %>%
  mutate(Title = str_to_title(Title)) 

# Format Fit Table
stage2_cmpk %>%
  gt() %>% 
  tab_options(column_labels.font.weight = "bold") %>%
  fmt_number(
    7,
    decimals = 2,
    drop_trailing_zeros = TRUE,
    suffixing = TRUE
  ) %>%
  fmt_number(c(3:6),
             decimals = 2) %>% 
    fmt_number(8,decimals = 2,
             drop_trailing_zeros=TRUE,
             suffixing = TRUE) %>% 
  fmt(8, fns = function(x) 
    ifelse(x>100, ">100",
           scales::number(x, accuracy = .1))) %>% 
  tab_style(
    style = list(
      cell_text(weight = "bold")
      ),
    locations = list(cells_body(
     columns = BIC,
     row = BIC == min(BIC[1:nrow(stage2_cmpk)]) 
    ),
    cells_body(
     columns = aBIC,
     row = aBIC == min(aBIC[1:nrow(stage2_cmpk)])
    ),
    cells_body(
     columns = CAIC,
     row = CAIC == min(CAIC[1:nrow(stage2_cmpk)])
    ),
    cells_body(
     columns = AWE,
     row = AWE == min(AWE[1:nrow(stage2_cmpk)])
    ),
    cells_body(
     columns = cmPk,
     row =  cmPk == max(cmPk[1:nrow(stage2_cmpk)])
     ),
    cells_body(
     columns = BF, 
     row =  BF > 10)
  )
)
```

Comparing two profiles
```{r}
a <- plotMixtures(output_pisa$model_2_class_2.out, ci = 0.95, bw = FALSE)

b <- plotMixtures(output_pisa$model_6_class_2.out, ci = 0.95, bw = FALSE)

a + labs(title = "Model 2 two-profile") + theme(plot.title = element_text(size = 12)) + b + labs(title = "Model 4 two-profile") +
    theme(plot.title = element_text(size = 12))
```

a initial look, not the final figure
```{r}
source(here("33-k-means", "functions", "plot_lpa.R"))  #file in the folder

plot_lpa(model_name = output_pisa$model_2_class_2.out)
```
another version: this one changed the color of two profiles and the variable names (all labels)
```{r}
plot_lpa <- function(model_name) {

  # Extract and reshape mean estimates by class
  pp_plots <- data.frame(model_name$parameters$unstandardized) %>%
    mutate(LatentClass = sub("^", "Class ", LatentClass)) %>%
    filter(paramHeader == "Means") %>%
    filter(LatentClass != "Class Categorical.Latent.Variables") %>%
    dplyr::select(est, LatentClass, param) %>%
    pivot_wider(names_from = LatentClass, values_from = est) %>%
    relocate(param, .after = last_col())
  
  # Extract class proportions
  c_size <- as.data.frame(model_name$class_counts$modelEstimated$proportion) %>%
    dplyr::rename("cs" = 1) %>%
    mutate(cs = round(cs * 100, 2))
  
  # Rename class labels with proportions
# Keep class labels without proportions
  colnames(pp_plots) <- c("Motivated but Unconfident", "Confident but Disengaged", "param")

  
  # Melt data into long format
  plot_data <- pp_plots %>%
    dplyr::rename("param" = ncol(pp_plots)) %>%
    reshape2::melt(id.vars = "param") %>%
    mutate(
     param = factor(param,
                   levels = c("Z_AMSC", "Z_ASCI_IA", "Z_CSC", "Z_EB", "Z_MSLQ"),
                   labels = c("Motivation", "Attitudes", "Self-concept", "Effort beliefs", "Self-efficacy")),
     variable = factor(variable)
    )
  
  # Title
  name <- str_to_title(model_name$input$title)
  
  # Define color palette (can customize further)
  #class_colors <- c("#1b1b1b", "#595959", "#a6a6a6", "#d9d9d9")  # Extend if >2 classes
  class_colors <- c("#F8766D", "#00BFC4")
  
  # Plot
  p <- plot_data %>%
    ggplot(
      aes(
        x = param,
        y = value,
        shape = variable,
        colour = variable,
        lty = variable,
        group = variable
      )
    ) +
    geom_point(size = 4) + 
    geom_line() +
    scale_x_discrete("") +
    scale_color_manual(values = class_colors[1:nlevels(plot_data$variable)]) +  # custom colors
    labs(title = glue("{name} Profile Plot"), y = "Means") +
    theme_cowplot() +
    theme(
      text = element_text(family = "serif", size = 12),
      legend.key.width = unit(.5, "line"),
      legend.text = element_text(family = "serif", size = 12),
      legend.title = element_blank(),
      legend.position = "top"
    )
  
  return(p)
}

plot_lpa(model_name = output_pisa$model_2_class_2.out)
```

Final figure

```{r}
plot_lpa <- function(model_name) {

  # Extract and reshape mean estimates by class
  pp_plots <- data.frame(model_name$parameters$unstandardized) %>%
    mutate(LatentClass = sub("^", "Class ", LatentClass)) %>%
    filter(paramHeader == "Means") %>%
    filter(LatentClass != "Class Categorical.Latent.Variables") %>%
    dplyr::select(est, LatentClass, param) %>%
    pivot_wider(names_from = LatentClass, values_from = est) %>%
    relocate(param, .after = last_col())
  
  # Extract class proportions
  c_size <- as.data.frame(model_name$class_counts$modelEstimated$proportion) %>%
    dplyr::rename("cs" = 1) %>%
    mutate(cs = round(cs * 100, 2))
  
  # Rename class labels with proportions
# Keep class labels without proportions
  colnames(pp_plots) <- c("Motivated but Unconfident", "Confident but Disengaged", "param")

  
  # Melt data into long format
  plot_data <- pp_plots %>%
    dplyr::rename("param" = ncol(pp_plots)) %>%
    reshape2::melt(id.vars = "param") %>%
    mutate(
     param = factor(param,
                   levels = c("Z_AMSC", "Z_ASCI_IA", "Z_CSC", "Z_EB", "Z_MSLQ"),
                   labels = c("Chemistry\nmotivation", "Chemistry\nattitudes", "Chemistry\nself-concept", "Effort beliefs", "Self-efficacy")),
     variable = factor(variable)
    )
  
  # Title
  name <- str_to_title(model_name$input$title)
  
  # Define color palette (can customize further)
  #class_colors <- c("#1b1b1b", "#595959", "#a6a6a6", "#d9d9d9")  # Extend if >2 classes
  class_colors <- c("#F8766D", "#005B96")  # red and dark blue
  
  # Plot
  p <- plot_data %>%
    ggplot(
      aes(
        x = param,
        y = value,
        shape = variable,
        colour = variable,
        lty = variable,
        group = variable
      )
    ) +
    geom_line(linewidth = 1) +
    geom_point(size = 4) +
    scale_x_discrete("") +
    scale_y_continuous(
    limits = c(-0.8, 0.8),
    breaks = c(-0.75, -0.5, -0.25, 0, 0.25, 0.5, 0.75)
    ) +
    scale_color_manual(values = class_colors[1:nlevels(plot_data$variable)]) +  # custom colors
    #labs(title = glue("{name} Profile Plot"), y = "Standardized indicator means") +
    labs(title = NULL, y = "Standardized indicator means") +
    theme_cowplot() +
    theme(
      text = element_text(family = "sans", size = 12),         # serif font
      axis.text = element_text(family = "sans", size = 12),    # axis labels
      axis.title = element_text(family = "sans", size = 14),    # axis labels
      legend.text = element_text(family = "sans", size = 12),  # legend text
      legend.title = element_text(family = "sans", size = 14),
      #legend.position = "top",
      legend.position = "none",
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
      axis.line = element_line(colour = "black", linewidth = 0.3)
    )
  
  return(p)
}

plot_lpa(model_name = output_pisa$model_2_class_2.out)

p <- plot_lpa(model_name = output_pisa$model_2_class_2.out)
ggsave("profile_plot.png", plot = p, width = 9, height = 5.6, units = "in", dpi = 600, bg = "white")

```


Three step ML 
```{r}
lpa_result <- pre_clean %>%
  dplyr::select(z_ASCI_IA, z_AMSC, z_EB, z_MSLQ, z_CSC) %>%
  estimate_profiles(n_profiles = 2, model = 2)  # Adjust the number of profiles as needed
```

```{r}
classified_data <- bind_cols(pre_clean, get_data(lpa_result))
```

```{r}
t.test(exam ~ Class, data = classified_data, var.equal = TRUE)
```
```{r}
classified_data %>%
  group_by(Class) %>%
  summarise(
    n = n(),
    mean = mean(exam, na.rm = TRUE),
    sd = sd(exam, na.rm = TRUE)
  )
```

wald test

```{r}
lm_model <- lm(exam ~ Class, data = classified_data)
summary(lm_model)
```

```{r}
Anova(lm_model, type = 3)  
```
Kmeans
```{r}
pre_clean <- pre_clean %>%
  mutate(across(all_of(c("ASCI_IA", "AMSC", "EB", "MSLQ", "CSC")), ~ scale(.)[,1], .names = "z_{.col}"))

dfk <- pre_clean %>%
  dplyr::select(z_ASCI_IA, z_AMSC, z_EB, z_MSLQ, z_CSC, exam, grading)
```

```{r}
# Step 1: Select specific variables for clustering
dfk_selected <- dfk %>%
  dplyr::select(z_AMSC, z_ASCI_IA, z_CSC, z_EB, z_MSLQ) 

# Step 2: Scale the selected data
dfk_scaled <- scale(dfk_selected)

# Step 3: Set random seed for reproducibility
set.seed(123)

# Step 4: Run K-means for k = 2 to 10 and store total within-cluster sum of squares
wss <- numeric(9)
kmeans_results <- list()

for (k in 2:10) {
  km <- kmeans(dfk_scaled, centers = k, nstart = 10, iter.max = 25)
  wss[k - 1] <- km$tot.withinss
  kmeans_results[[as.character(k)]] <- km
}

# Step 5: Plot Elbow Method to choose optimal k
plot(2:10, wss, type = "b", pch = 19,
     xlab = "Number of Clusters (k)",
     ylab = "Total Within-Cluster Sum of Squares",
     main = "Elbow Method for Optimal k")

# Step 6: Assign cluster labels to the original selected data for chosen k
chosen_k <- 2
dfk_clustered <- dfk_selected %>%
  mutate(Cluster = factor(kmeans_results[[as.character(chosen_k)]]$cluster))

# Step 7: Visualize the clusters using PCA
fviz_cluster(kmeans_results[[as.character(chosen_k)]], data = dfk_scaled,
             geom = "point", ellipse.type = "convex",
             main = paste("K-means Clustering (k =", chosen_k, ")"))
```
```{r}
kmeans_results[["2"]]$size
```

```{r}
fviz_nbclust(dfk_clustered, kmeans, method = "wss") #choose 3
```

elbow plot
```{r}
# Step 1: Add WSS for k = 1
# (Only if it wasn’t previously computed)
set.seed(123)
km1 <- kmeans(dfk_scaled, centers = 1, nstart = 10, iter.max = 25)
wss_full <- c(km1$tot.withinss, wss)  # Add k=1 to the front

# Step 2: Recreate full dataframe
elbow_df <- data.frame(
  k = 1:10,
  wss = wss_full
)

# Step 3: Create the updated plot
plot.wss <- ggplot(elbow_df, aes(x = k, y = wss)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(limits = c(400, 1400)) +  # << Set y-axis range
  labs(
    x = expression("Number of clusters (" * italic(k) * ")"),
    y = "Total Within Sum of Squares (WSS)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    axis.text = element_text(size = 12, color = "black"),
    axis.title = element_text(size = 14, color = "black"),
    axis.ticks.y = element_line(color = "black")
  )

plot.wss

ggsave("cluster-wss.png", plot = plot.wss, width = 9, height = 5.6, units = "in", dpi = 600, bg = "white")
```


```{r}
fviz_nbclust(dfk_clustered, kmeans, method = "silhouette") #choose 2

# Generate silhouette plot and customize it
plot.silhouette <- fviz_nbclust(dfk_clustered, kmeans, method = "silhouette") +
  geom_line(color = "steelblue", linewidth = 3) +
  geom_point(color = "steelblue", size = 3) +
  labs(
    x = expression("Number of clusters (" * italic(k) * ")"),
    y = "Average silhouette width",
    title = NULL
  ) +
  scale_x_discrete() +  # ← key fix here
  scale_y_continuous(limits = c(0, 0.3)) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    axis.text = element_text(size = 12, color = "black"),
    axis.title = element_text(size = 14, color = "black"),
    axis.ticks.y = element_line(color = "black")
  )

plot.silhouette

ggsave("silhouette.png", plot = plot.silhouette, width = 9, height = 5.6, units = "in", dpi = 600, bg = "white")
```
```{r}
# Reshape the cluster centers data for plotting
long_data <- kmeans_results[["2"]]$centers %>%
  as.data.frame() %>%
  mutate(Cluster = factor(rownames(.))) %>%
  pivot_longer(-Cluster, names_to = "Variable", values_to = "Mean")

long_data$Cluster <- dplyr::recode(long_data$Cluster,
                            "1" = "Motivated but Unconfident",
                            "2" = "Confident but Disengaged")

# Define readable labels for each variable
variable_labels <- c("z_AMSC" = "Motivation",
                     "z_ASCI_IA" = "Attitudes",
                     "z_CSC" = "Self-concept",
                     "z_EB" = "Effort beliefs",
                     "z_MSLQ" = "Self-efficacy")

# Plot the cluster profiles
ggplot(long_data, aes(x = Variable, y = Mean, group = Cluster, color = Cluster)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Cluster Profiles on Key Academic Variables",
       y = "Cluster Mean (Z-score)",
       x = NULL) +
  scale_x_discrete(labels = variable_labels) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),  # Horizontal text
    plot.title = element_text(hjust = 0.5),
    legend.position = "top",
    legend.direction = "horizontal",
    panel.grid.major = element_blank(),  # remove major grid lines
    panel.grid.minor = element_blank(),   # remove minor grid lines
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black")
  )
```

final cluster plot
```{r}
# Define consistent color palette
cluster_colors <- c("Motivated but Unconfident" = "#F8766D",
                    "Confident but Disengaged" = "#005B96")

# Updated variable labels with line breaks
variable_labels <- c(
  "z_AMSC"    = "Chemistry\nmotivation",
  "z_ASCI_IA" = "Chemistry\nattitudes",
  "z_CSC"     = "Chemistry\nself-concept",
  "z_EB"      = "Effort beliefs",
  "z_MSLQ"    = "Self-efficacy"
)

# Generate the plot
plot.cluster <- ggplot(long_data, aes(x = Variable, y = Mean, group = Cluster, color = Cluster, linetype = Cluster, shape = Cluster)) +
  geom_line(linewidth = 1) +
  geom_point(size = 4) +
  scale_color_manual(values = cluster_colors) +
  scale_x_discrete(labels = variable_labels) +
  scale_y_continuous(
    limits = c(-0.9, 0.9),
    breaks = c(-0.75, -0.5, -0.25, 0, 0.25, 0.5, 0.75)
  ) +
  labs(
    #title = "Cluster Profiles on Key Academic Variables",
    y = "Standardized indicator means",
    x = NULL
  ) +
  theme_classic(base_family = "sans") +
  theme(
    text = element_text(size = 12),
    axis.text = element_text(size = 12, colour = "black"),
    axis.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    legend.title = element_blank(),
    legend.position = "none",  # or "top" if you want to keep it
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.8),
    axis.line = element_line(colour = "black", linewidth = 0.3)
  )

plot.cluster

ggsave("cluster.png", plot = plot.cluster, width = 9, height = 5.6, units = "in", dpi = 600, bg = "white")

```



auxiliary
```{r}
dfk_clustered <- dfk %>%
  dplyr::select(z_ASCI_IA, z_AMSC, z_EB, z_MSLQ, z_CSC, exam, grading) %>%  # include your two auxiliary variables
  drop_na() %>%  # drop rows with missing values
  mutate(Cluster = factor(kmeans_results[["2"]]$cluster))
```

```{r}
# t-test for exam by Cluster (equal variances)
t.test(exam ~ Cluster, data = dfk_clustered, var.equal = TRUE)
```

```{r}
# t-test for grading by Cluster (equal variances)
t.test(grading ~ Cluster, data = dfk_clustered, var.equal = TRUE)
```

```{r}
dfk_clustered %>%
  group_by(Cluster) %>%
  summarise(
    Mean = mean(exam, na.rm = TRUE),
    SD = sd(exam, na.rm = TRUE),
    N = n()
  )
```

```{r}
df_combined <- cbind(dfk_clustered, classified_data %>% rename(exam_2 = exam, grading_2 = grading))
```

```{r}
tab <- df_combined %>%
  count(Cluster, Class) %>%
  group_by(Class) %>%
  mutate(percent = round(100 * n / sum(n), 1)) %>%
  ungroup()

tab <- tab %>%
  group_by(Class) %>%
  mutate(percent = round(100 * n / sum(n), 1)) %>%
  ungroup()

print(names(tab))
```
```{r}
tab %>%
  gt() %>%
  tab_header(
    title = "Contingency Table: K-means Cluster × LPA Class"
  ) %>%
  cols_label(
    Cluster = "K-means Cluster",
    Class = "LPA Class",
    n = "Count",
    percent = "Percent (%)"
  ) %>%
  fmt_number(columns = percent, decimals = 1)
```

