---
title: "Connecting Math Attitudes With STEM Career Attainment: A Latent Class Analysis Approach" 
author: "Dina Arch"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(haven)
library(tidyverse)
library(sjPlot)
library(gt)
library(janitor)
library(modelsummary)
library(here)
library(glue)
library(MplusAutomation)
library(patchwork)
library(reshape2)
library(cowplot)
library(naniar)

#install.packages("devtools")
#library(devtools)

#install_github("michaelhallquist/MplusAutomation")
here::i_am("dang_analysis.Rmd")
```

# Education Longitudinal Study

-   BY = Base Year (2002) - 10th grade

    -   Completed the baseline survey of high school sophomores in spring term 2002.

-   F1 = First follow up (2004) - 12th grade

    -   Most sample members were seniors, but some were dropouts or in other grades (early graduates or retained in an earlier grade).

-   F2 = Second follow up (2006) - Post-high-school follow-up

-   F3 = Third follow up (2012) - Kids are ~26yo

## Item Descriptions

### Base Year Covariates

*English Proficiency*
BYS70A: How well student understands English (1 = Very well, 4 = Not at all) 
BYS70B: How well student speaks English (1 = Very well, 4 = Not at all) 
BYS70C: How well student reads English (1 = Very well, 4 = Not at all)
BYS70D: How well 10th-grader writes English (1 = Very well, 4 = Not at all) 
BYTM12B: Student behind on schoolwork due to limited proficiency in English
BYSTLANG: English if student's native language

*Demographics*
BYSEX: 1 = Male, 2 = Female
BYRACE_R: All -5 since restricted
BYSES2QU: SES Quartile

### Base Year Indicators

BYS87A: When I do mathematics, I sometimes get totally absorbed
BYS87C: Because doing mathematics is fun, I wouldn't want to give it up
BYS87F: Mathematics is important to me personally

BYS89A: I'm confident that I can do an excellent job on my math tests
BYS89B: I'm certain I can understand the most difficult material presented in math texts
BYS89L: I'm confident I can understand the most complex material presented by my math teacher
BYS89R: I'm confident I can do an excellent job on my math assignments
BYS89U: I'm certain I can master the skills being taught in my math class

*Math Attitude BYS87 Items*

[Description]{.underline}: How much do you agree or disagree with the following
statements?

[Response options]{.underline}:

1 - Strongly agree

2 - Agree

3 - Disagree

4-  Strongly disagree


*Math Self-Efficacy BYS89/F1S18 Items*

[Description]{.underline}: How much do you agree or disagree with the following
statements?

[Response options]{.underline}:

1 - Almost never

2 - Sometimes

3 - Often

4-  Almost always

### F1 Indicators

F1S18A: I'm confident that I can do an excellent job on my math tests
F1S18B: I'm certain I can understand the most difficult material presented in my math textbooks
F1S18C: I'm confident I can understand the most complex material presented by my math teacher
F1S18D: I'm confident I can do an excellent job on my math assignments
F1S18E: I'm certain I can master the skills being taught in my math class

### Outcomes

F3ONET2CURR: 2 digit code for most recent job
F3ONET6CURR: 6 digit code for most recent job
F3BYTSCWT: Panel weight

## Read in data

Read in ELS `.sav` file:

```{r, eval = FALSE}
els_sav <- read_sav(here("els_data", "els_02_12_byf3pststu_v1_0.sav")) 
```

View codebook:

```{r, eval = FALSE}
#view_df(els_sav)
```

Save subset:

```{r, eval = FALSE}
els_subset <- els_sav %>% 
  clean_names() %>%
  dplyr::select(
    stu_id, strat_id, psu,
    bys87a,bys87c,bys87f,bys89a,bys89b,bys89l,bys89r,bys89u,
    f1s18a,f1s18b,f1s18c,f1s18d,f1s18e,
    bys70a,bys70b,bys70c,bys70d,bystlang,bytm12b,bysex,byrace_r,
    byses2qu,f3onet2curr,f3onet6curr, f3bystemoc30, f3bytscwt
         ) %>%
  mutate(school = paste0(strat_id,psu))

view_df(els_subset)

#write_csv(els_subset, here("els_data", "els_dang_analysis.csv"))
```

Read in ELS data file, `els_dang_analysis.csv`.

```{r, eval=TRUE}
bys87_items <- c("bys87a","bys87c","bys87f")

f1s18_items <- c("f1s18a","f1s18b","f1s18c","f1s18d","f1s18e")

bys89_items <- c("bys89a","bys89b","bys89l","bys89r","bys89u")

els_data <- read_csv(
  here("els_data", "els_dang_analysis.csv"),
  na = c("-9", "-8", "-6", "-5", "-4", "-7", "-1", "-3", "-2", "-99")
) %>%
  # Dichotomize English proficiency
  mutate(across(
    c(bys70a, bys70b, bys70c, bys70d),
    ~ case_when(. %in% c(1, 2) ~ 1, 
                . %in% c(3, 4) ~ 0, 
                TRUE ~ NA_real_)
  )) %>%
  mutate(across(
    c(f3onet2curr),
    ~ case_when(is.na(.) ~ NA_real_,
                . %in% c(11, 15, 17, 19, 25, 45, 51) ~ 1, 
                TRUE ~ 0)
  )) %>%
  mutate(female = case_match(bysex, 
                             2 ~ 1, 
                             1 ~ 0, 
                             .default = NA_real_)) %>%
  mutate(ses_dichotomized = case_match(
    byses2qu,
    1 ~ 1,
    NA ~ NA,
    .default = 0   
  )) %>%
  mutate(across(
    all_of(bys87_items),
    ~ case_when(. %in% c(1, 2) ~ 1,
                . %in% c(3, 4) ~ 0,
                TRUE ~ NA_real_)
  )) %>%
  mutate(across(all_of(c(
    bys89_items, f1s18_items
  )), ~ case_when(
    . %in% c(1, 2) ~ 0,
    . %in% c(3, 4) ~ 1, 
    TRUE ~ NA_real_
  )))

summary(els_data)

summary(factor(els_data$bys70a))
summary(factor(els_data$bys70b))
summary(factor(els_data$bys70c))
summary(factor(els_data$bys70d))
summary(factor(els_data$bystlang))
summary(factor(els_data$f3onet2curr))
summary(factor(els_data$female))
summary(factor(els_data$bysex))
summary(factor(els_data$byses2qu))
summary(factor(els_data$ses_dichotomized))
```

---

# Class Enumeration Using PSEM

Experiment with different penalty hyperparameters

```{r, eval=FALSE}
if (!dir.exists(here("psem","penalty_enum"))) {
  dir.create(here("psem","penalty_enum"), recursive = TRUE)
}

penalty_enum <- lapply(seq(from=0.1,to=1.5,length.out=8),function(k) {
  m_psem_enum <- mplusObject(
        
      TITLE = glue("PSEM ALF({k})"), 
    
      VARIABLE =
      "usev = bys87a-bys89u bys70a-bystlang;
      categorical = bys87a-bys89u;
      classes = c(4);
      
      WEIGHT = f3bytscwt;",
    
    ANALYSIS = 
      "type = mixture;
      estimator = mlr;
      Starts = 800 200;
      proc = 12;",
    
    MODEL = 
    "%OVERALL% 
    c ON bys70a-bystlang;
    ! Direct effects:
    bys87a-bys89u ON bys70a-bystlang (d1-d64);",
    
    MODELPRIORS = 
    glue("d1-d64~ALF(0,{k});"),
    
    OUTPUT = "sampstat residual tech11 tech14;",
  
    usevariables = colnames(data_binary),
    rdata = data_binary)
  
    step1_fit <- mplusModeler(m_psem_enum, 
                              dataout=glue(here("psem","penalty_enum","penalty_enum.dat")),
                              modelout=glue(here("psem","penalty_enum","penalty_enum_{k}.inp")) ,
                              check=TRUE, run = TRUE, hashfilename = FALSE)
})
```

* Since we used MLR there is a correction that needs to be done to the likelihood, at least for Chi-Square tests. Not sure if this also applies to subjective comparison of log-likelihood as done below (https://www.statmodel.com/discussion/messages/11/8230.html?1590018965)

* For some reason ALF(0, 0) gives all NaN instead of just 0 for all the direct effects

```{r}
psem_penalty <- readModels(here("psem","penalty_enum"), quiet = TRUE)

ll_values <- sapply(psem_penalty, function(x) x$summaries$LL)
sig_effects <- sapply(psem_penalty, function(x) sum(x$parameters$unstandardized$pval < 0.05))

sig_dir_effects <- sapply(psem_penalty, function(x) {
  params <- x$parameters$unstandardized
  # Keep pvals that are NOT thresholds or intercepts AND do NOT contain "C#" (should just be the direct effects)
  keep <- !(params$paramHeader %in% c("Thresholds", "Intercepts")) & 
          !grepl("C#", params$paramHeader)
  sum(params$pval[keep] < 0.05, na.rm = TRUE)
})

df <- data.frame(
  model_name = names(ll_values),
  log_likelihood = as.numeric(ll_values),
  effects = sig_effects,
  dir_effects = sig_dir_effects
)

df$penalty_value <- as.numeric(sub(".*_([0-9.]+)\\.out", "\\1", df$model_name))
```

```{r}
ggplot(data=df, mapping=aes(x=ordered(penalty_value),y=dir_effects)) +
  geom_point() +
  geom_line() +
  labs(x = "Penalty (Alignment Function Variance)",
       y = "Significant Direct Effects")
```

Estimation is the same for all classes, so all these numbers are divided by 4 in terms of a path diagram.

Interesting that a high penalty decreases the total amount of significant direct effects. I'm really not sure what this happens. 

```{r}
ggplot(data=df, mapping=aes(x=ordered(penalty_value),y=log_likelihood)) +
  geom_point() +
  geom_line() +
  labs(x = "Penalty (Alignment Function Variance)",
       y = "Significant Direct Effects")
```

I would say ALF(0,0.7) is ideal because it provides a simple model with very good log likelihood, but there's definitely an argument to keep all the direct effects in the model.

Now Muthen hardcodes just the significant effects. I'm Not sure if this is the best approach because it ignores potential interactions with non-significant effects, but I'll continue with his procedure.

```{r, eval=FALSE}
m_psem_enum <- mplusObject(
      
    TITLE = "8 Us, Hardcoded Dir Effects", 
  
    VARIABLE =
    "usev = bys87a-bys89u bys70a-bystlang;
    categorical = bys87a-bys89u;
    classes = c(4);
    
    WEIGHT = f3bytscwt;",
  
  ANALYSIS = 
    "type = mixture;
    estimator = mlr;
    Starts = 800 200;
    proc = 12;",
  
  MODEL = 
  "%OVERALL% 
  c ON bys70a-bystlang;
  ! Direct effects:
  BYS87A on BYS70B BYS70D;
  BYS89R on BYS70B BYSEX;",
  
  OUTPUT = "sampstat residual tech11 tech14;",

  usevariables = colnames(data_binary),
  rdata = data_binary)

psem_enum_fit <- mplusModeler(m_psem_enum, 
 dataout=here("psem", "dir_effects.dat"),
 modelout=here("psem", "dir_effects.inp"),
 check=TRUE, run = TRUE, hashfilename = FALSE)
```

```{r}
psem <- readModels(here("psem"), quiet = TRUE)
plot_lca(psem)
```

There are major difference in class enumeration once direct effects are modelled

* Class 4 is basically the same

* Class 3 has similar but has much higher rates of endorsing the last 3 items

* Classes 1 is similar but much lower rate of endorsing BYS87C

* Class 2 is completely different
